---
title: "Drawing a mosaic plot with `vcd::mosaic()`"
author: "Joyce Robbins"
date: "2/15/2018"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, fig.height = 3.5,
                      fig.width = 5, fig.align = 'center',
                      warning = FALSE, message = FALSE,
                      cache = TRUE)
```

## The Data

```{r}
df <- read.csv("../data/MusicIcecream.csv")
df
```

### Order of splits

It is best to draw mosaic plots incrementally: start with splitting on one variable and then add additional variables one at a time.  The full mosaic plot will have one split per variable.

Important: if your data is in a data frame (see above), **the count column must be called `Freq`**. (Other data structure options are tables, matrices -- for 2 variables -- or objects of class `structable` -- see `>?vcd::structable`.)

Also note that all of these plots are drawn with **`vcd::mosaic()`** *not `mosaicplot()`*.

Split on `Age` only:

```{r}
library(vcd)
mosaic(~Age, df)
```

Split on `Age`, then `Music`:

```{r}
mosaic(Music ~ Age, df)
```

Note that the *first* split is between "young" and "old", while the second set of splits divides each age group into "classical" and "rock".

Split on `Age`, then `Music`, then `Favorite`:

```{r}
mosaic(Favorite ~ Age + Music, df)
```

### Direction of splits

Note that in the previous example, the direction of the splits is as follows:

1. `Age` -- horizontal split

2. `Music` -- vertical split

3. `Favorite` -- horizontal split

This is the default direction pattern: alternating directions beginning with horizontal. Therefore we get the same plot with the following:

```{r}
mosaic(Favorite ~ Age + Music, 
       direction = c("h", "v", "h"), df)
```

The directions can be altered as desired.  For example, to create a doubledecker plot, make all splits vertical except the last one:

```{r}
mosaic(Favorite ~ Age + Music,
       direction = c("v", "v", "h"), df)
```

Note that the direction vector is in order of splits (`Age`, `Music`, `Favorite`), not in the order in which the variables appear in the formula, where the last variable to be split is listed first, before the "~".

### Options

#### Fill color:

```{r}
library(grid) # needed for gpar
mosaic(Favorite ~ Age + Music, 
       gp = gpar(fill = c("pink", "tan3")),
       df)
```

Note that the fill colors are applied to the dependent variable factor levels ("coffee" and "bubble gum" in this case.) 

#### Rotate labels:

```{r}
mosaic(Favorite ~ Age + Music, 
       labeling = labeling_border(rot_labels = c(45, -45, 0, 0)),
       df)
```

The `rot_labels = ` vector sets the rotation in degrees on the four sides of the plot in this order: *top, right, bottom, left*. (Different from the typical base graphics order!) The default is `rot_labels = c(0, 90, 0, 90)`.

#### Abbreviate labels:

```{r}
mosaic(Favorite ~ Age + Music, 
       labeling = labeling_border(abbreviate_labs = c(3, 1, 6)), 
       df)
```

Labels are abbreviated in the order of the splits (as for `direction = `). The abbreviation algorithm appears to return the specified number of characters after vowels are eliminated (if necessary).

For more formatting options, see `>?vcd::labeling_border`.

#### Remove spacing between cells

```{r}
mosaic(Favorite ~ Age + Music,
       spacing = spacing_equal(sp = unit(0, "lines")),
       df)
```

For more details, see `>?vcd::spacings`

#### Change border color (must also set fill(?))

```{r}
mosaic(Favorite ~ Age + Music,
       gp = gpar(fill = c("lightblue", "blue"),
                 col = "white"),
       spacing = spacing_equal(sp = unit(0, "lines")),
       df)
```

Please feel free to suggest changes by submitting a pull request or ask questions / offer comments by opening an issue.
